Regularized target encoding outperforms traditional methods in supervised machine learning with high cardinality features

### 2021-09-09 publication/manuscript.Rmd
shorten text

abstract: |
  Because most machine learning (ML) algorithms are designed for numerical inputs, efficiently encoding categorical variables is a crucial aspect during data analysis. An often encountered problem are high cardinality features, i.e. unordered categorical predictor variables with a high number of levels. We study techniques that yield numeric representations of categorical variables which can then be used in subsequent ML applications. We focus on the impact of those techniques on a subsequent algorithm's predictive performance, and -- if possible -- derive best practices on when to use which technique. We conducted a large-scale benchmark experiment, where we compared different encoding strategies together with five ML algorithms (lasso, random forest, gradient boosting, k-nearest neighbours, support vector machine) using datasets from regression, binary- and multiclass- classification settings. Throughout our study, regularized versions of target encoding (i.e. using target predictions based on the feature levels in the training set as a new numerical feature) consistently provided the best results. Traditional encodings that make unreasonable assumptions to map levels to integers (e.g. integer encoding) or to reduce the number of levels (possibly based on target information, e.g. leaf encoding) before creating binary indicator variables (one-hot or dummy encoding) were not as effective.
keywords:
- supervised machine learning
- benchmark
- high-cardinality categorical features
- target encoding
- dummy encoding
- generalized linear mixed models

@@

*Target-based* methods try to incorporate information about the target values associated with a given level.
Early strategies aimed to reduce the number of levels by methods like hierarchical clustering or decision trees based on statistics of the target variable, although this has been rarely described in the scientific literature [for a brief mention, see @micci_barreca_2001].
The basic idea of more advanced methods called target, impact, mean, or likelihood encoding is to use the training set to make a simple prediction of the target for each level of the categorical feature, and to use the prediction as the numerical feature value $\hat{x}_l$ for the respective level.
An early formal description of this strategy is @micci_barreca_2001.
Similar approaches are called *weights of evidence* in the credit scoring literature @hand_1997.
In simple target encoding for regression problems, the mean target value in the training set from all observations with a certain feature level is used as the numeric value to encode that level for all observations: $\hat{x}_l = \frac{\sum_{i:x^{train}_i = l}y^{train}_i}{N_l}$.
Simple target encoding often does not perform well with rare levels, where it tends to overfit to the training data and fails to generalize well for new observations.
In the extreme case of a categorical feature with unique values (e.g. some hashed ID variable) studied in @prokhorenkova_2018, the mean target for each level of this feature in simple target encoding is similar to the true target value of a single observation.
Based on the encoded feature, all observations can be predicted perfectly in the training set, even if the original variable did not contain any useful information.
ML models would place a high priority on such an encoded feature during training but would perform badly on test data.
To avoid this, practitioners often use regularized target encoding by adding a smoothing parameter $\epsilon$ which shrinks those effects towards the global mean [@micci_barreca_2001].
An alternative strategy is to combine target encoding with cross-validation (CV) techniques [@prokhorenkova_2018].

@@

Several small-scale studies have been previously conducted.
Those works did not yield conclusive results due to narrower scopes or not considering high cardinality variables.
One benchmark ($6$ datasets) on encoding high cardinality features (cardinality between $103$ and $9095$) in combination with gradient boosting has been published on the Kaggle forums [@prokopev_2018].
Different versions of target encoding are compared with indicator, integer, and frequency encoding.
They recommend combining the smoothed version of target encoding with 4- or 5-fold CV, never using simple target encoding and using indicator encoding only for small datasets.
Interestingly, frequency encoding did perform well in many cases.
@coors_2018 performed a benchmark ($12$ datasets) on encoding high cardinality features (maximum number of levels per dataset between 10 and 25847) when developing the automatic gradient boosting (autoxgboost) library [@thomas_2018]. They compared different variants of target encoding with integer and indicator encoding. In their benchmarks, target encoding only improved over target-agnostic methods on $2$ datasets, while it led to worse results on 4 datasets. For a smaller number of levels, indicator and integer encoding yielded similar results.
As those studies only consider gradient boosting and a limited amount of datasets, it is unclear whether results generalize to other datasets and ML algorithms.
Several other publications studied encoding text data based on similarity [@cerda_2018; @cerda_2020] and employed indicator or target encoding as baselines.
Another line of work studies variable encodings employed within specific ML models.
@wright_2019 study treatments for categorical variables in random forests together with dummy and integer encoding (18 datasets, cardinality between $3$ and $38$), concluding that indicator and integer encoding perform subpar in comparison to methods that re-order levels according to the target variable.
@prokhorenkova_2018 compared their new CatBoost variant of target encoding to smoothed target encoding with no CV, hold-out, and leave-one-out CV on $8$ datasets.
While the CatBoost method performed best, hold-out came second and target encoding without CV performed worst.
An overview of encoding techniques tailored to be used together with Neural Networks is provided in @hancock.
They present a survey of indicator-based and embedding-based methods but do not conduct a benchmark study.
One widely adopted approach, in particular, has been proposed by @guo_2016 under the term *entity embeddings*.
In contrast, our work focuses on high cardinality variables and techniques that are agnostic to the subsequent ML method, which has not been done before.
We furthermore study this problem on a sufficiently large variety of datasets and settings to allow for more compelling conclusions.
\newline

The main goal of our study is to assess the impact different categorical encoding techniques have on subsequent models' predictive performance.
As the optimal encoding might differ depending on the ML algorithm, we consider various state-of-the-art algorithms, *regularized linear models*, *random forests*, *gradient tree boosting*, *k-nearest neighbors*, and *support vector machines*.
To find default settings for high cardinality features, we analyze a variety of datasets with different characteristics including regression, binary classification and multiclass classification problems.
As our methods vary in runtime and complexity, we aim to learn whether more complex methods, in general, are to be preferred, or if simpler approaches suffice.
Furthermore, the relationship between a feature's cardinality and the choice of encoding technique has not been investigated, which we study by varying the minimum number of levels above which features are transformed.

@@

We first survey various categorical encoding techniques and conduct a comprehensive benchmark study of a broad set of categorical encoding techniques with a focus on high cardinality features.
We study $7$ different encoding techniques in conjunction with $5$ commonly used ML algorithms across $24$ datasets, both from a classification and a regression regime. We give a detailed description of our study design to highlight important considerations for studying high cardinality features.
Our results provide an overview of the performance of various approaches heavily used in the literature. After a discussion of results concerning predictive performance, we provide further analyses seeking to inform practitioners which methods to apply. This includes an important discussion on runtimes.

@@

Pseudocode is presented in the main text (for non-standard encodings), or the Supplementary Material (for standard methods).
An important detail is how encoding techniques treat new levels during the prediction phase, as this is often not obvious.

@@

The simplest strategy for categorical features is integer encoding, which is also often referred to as ordinal encoding.
To make sure that the order of the observed levels does not affect our results, we randomly map the observed levels from the training set to the integers $1$ to $L$.
Although new levels could be mapped to $L+1$ or $0$, model predictions would be completely arbitrary as the order of the integers does not contain any information.
Thus, we encode new levels as missing values and use mode imputation to obtain the integer which matches the most frequent level in the training set.
Integer encoding should be an acceptable strategy for tree-based models, which can separate all original levels with repeated splits, but less so for other algorithms.

@@

Frequency encoding simply maps each level to its observed frequency in the training set ($\hat{x}_l = N_l$).
On the one hand, this assumes a functional relationship between the frequency of a level and the target.
On the other hand, it implicitly reduces the number of levels, as the subsequent model can best differentiate between levels with dissimilar frequencies. This approach is typically used e.g. in natural language processing to encode token or n-gram counts.
We encode new levels with a frequency of $1$, although choosing the minimal observed frequency for any level during training might be another reasonable strategy.

@@

We use indicator encoding as an umbrella term for the two most common strategies to encode categorical features with a small to moderate number of levels: one-hot and dummy encoding.
One-hot encoding transforms the original feature into $L$ binary indicator columns, each representing one original level.
An observation is coded with $1$ for the indicator column representing its level ($x^{train}_i = l$) and $0$ for all other indicators.
Dummy encoding results in only $L-1$ indicator columns.
A reference feature level is chosen that is encoded with $0$ in all indicator columns.
For one-hot encoding, the zero vector can be used to encode new levels which were not observed during training.
For dummy encoding, this is not useful as it collapses new levels to the (often arbitrary) reference category.
In our study, the first level in alphabetical order is used as the reference category for dummy encoding.
We replace new levels in the prediction phase with the most frequent level in the training set.
In datasets with a high total number of feature levels, constructing all indicator variables can be detrimental to the predictive performance or at least increase the computational load.
As indicator encoding is practically infeasible for high cardinality variables (a feature with $10^5$ levels would add $10^5$ new columns), we limit the number of indicator coded features by collapsing rare levels beyond a varying threshold to a single *other* category before encoding.

@@

Hash Encoding can be used to compute indicator variables based on a hash function [@weinberger_2009].
The basic idea is to transform each feature level $l$ into an integer $hash(l) \in \mathbb{N}$, based on its label.
This integer is then transformed into an indicator representation, with $1$ in indicator column number $(hash(l) \mod hash.size) + 1$ and $0$ in all remaining columns [@kuhn_2019].
Some levels will be hashed to the same indicator representation.
The smaller the $hash.size$, the higher the number of collapsed levels.
The number of indicators is often effectively lower than $hash.size$, as some indicators can be constant in the training set (we remove those columns for both training and prediction).
Although hashing would allow us to jointly hash multiple features, we hash each feature separately to improve comparability with the other encoders.

@@

In leaf encoding, a decision tree is fitted on the training set to predict the target based on the categorical feature.
Each level is encoded by the number of the terminal node, in which an observation with the respective level ends up.
In that way, leaf encoding combines feature levels with similar target values.
We use the rpart package in R [@therneau_2018] which grows CARTs with categorical feature support that can be pruned based on internal performance estimates from 10-fold CV.
Thus, our leaf encoder automatically uses an "optimal" number of new levels.
To speed up the computation for multiclass classification, our implementation uses the ordering approach presented in @wright_2019.
New levels are encoded with the arbitrary number of the terminal node with most observations during training.
Encoded values are treated as a new categorical feature and encoded by one-hot encoding.
Our leaf encoder can be thought of as a simplification of the approach suggested by @grabczewski_2003.

@@

The first formal description of a so-called impact or target encoder was provided by @micci_barreca_2001.
The basic idea is to encode each feature level with the conditional target mean (regression) or the conditional relative frequency of one or more target classes (classification).
The impact encoder for classification uses a logit link and transforms the original feature into $C$ numeric features, each representing one target class.
A smoothing parameter $\epsilon$ is introduced to avoid division by zero.
This parameter could be further used to regularize towards the unconditional mean.
To allow comparing regularized (introduced below) with simple target encoding, we always choose a small $\epsilon = 0.0001$ for the impact encoder.
Similar techniques have been proposed in the literature:
Weights of evidence encoding from the credit scoring classification literature [@hand_1997] is almost identical to impact encoding, but does not
consider any regularization or centering.
Impact encoding is also referred to as a target or James-Stein encoding.

@@

Smoothed target encoding [@micci_barreca_2001] can be interpreted as a simple (generalized) linear mixed model (glmm) in which the target is predicted by a random intercept for each feature level in addition to a fixed global intercept.
This connection is described in @kuhn_2019.
To achieve regularized impact encoding, we constructed our own glmm encoders for regression, binary, and multiclass classification which are described in Algorithm \ref{alg:glmm-regr} (regression) and in the Supplementary Material (classification).
The encoded value for each level is based on the spherical conditional mode estimates.
In regression, the conditional modes are similar to the mean target value for each level, weighted by the relative observed frequency of that level in the training set [@gelman_2006]. For technical details, see @bates_2018.
Additionally, the estimate of the fixed intercept can be used during the prediction phase to encode new feature levels that were not observed in the training set.
In multiclass classification, we fit $C$ one vs. rest glmms resulting in one encoded feature per class.
An important advantage of using a glmm over impact encoding with a smoothing parameter is that a reasonable amount of regularization is determined automatically and tuning the complete ML pipeline is not necessary.

A further strategy to avoid overfitting in target encoding is to combine it with CV to train the encoder on independent observations without limiting the data to train the ML model.
To our knowledge, we provide the first implementation which combines target encoding based on glmms with CV.
During the training phase, we partition the data using CV into $n.folds$ and fit a glmm on each resulting training set.
For each observation, there is exactly one glmm that did not use that observation for model fitting and can be safely used for encoding.
Note that the $n.folds$ CV models (for $n.folds$ > 1) are only used during the training phase.
In the prediction phase, feature values are always encoded by a single glmm fitted to the complete training set.
We study this method in three different settings: without CV (noCV), with $5-$ (5CV) and with $10-$ fold CV (10CV).
In our study, we use the lmer (regression) and glmer (classification) functions from the lme4 package in R [@bates_2015] as an efficient way to fit glmms.
\begin{algorithm}
  \caption{GLMM Encoding Regression}
  \label{alg:glmm-regr}
  \begin{algorithmic}
    \State \underline{Training:} require $n.folds \in \mathbb{N}$
     \State fit simple random intercept model: $y^{train}_i = \beta_{0l} + \epsilon_{i} = \gamma_{oo} + u_l + \epsilon_{i}$ on $\mathcal{D}^{train}$
     \State with $u_l \overset{iid}{\sim} N(0,\tau^2)$, $\epsilon_i \overset{iid}{\sim} N(0,\sigma^2)$ and $x^{train}_i = l$, $l \in \mathcal{L}^{train}$
    \If{$n.folds = 1$}
      \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}
        \State $\hat{x}^{train}_i = \hat{\beta}_{0l}^{\mathcal{D}^{train}}$ with $x^{train}_i = l$
      \EndFor
    \Else\ {use $n.folds$ cross-validation scheme to make training sets $\mathcal{D}^{train}_1, \dots, \mathcal{D}^{train}_{n.folds}$}
      \State and fit simple random intercept model on each $\mathcal{D}^{train}_{m}$
      \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}
        \State $\hat{x}^{train}_i = \hat{\beta}_{0l}^{\mathcal{D}^{train}}$ with $x^{train}_i = l$ based on the model m
        \State with $(x^{train}_i, y^{train}_i) \notin \mathcal{D}^{train}_m$
      \EndFor
    \EndIf
    \State \underline{Prediction:}
    \For{$x^{new}$}
      \If{$x^{new} \in \mathcal{L}^{train}$}
        \State $\hat{x}^{new} = \hat{\beta}_{0l}^{\mathcal{D}^{train}}$ with $x^{new} = l$ based on full model fitted on $\mathcal{D}^{train}$
      \Else\ {$\hat{x}^{new} = \hat{\gamma}_{00}$ based on full model fitted on $\mathcal{D}^{train}$}
      \EndIf
    \EndFor
  \end{algorithmic}
\end{algorithm}

@@

We include three control conditions to better understand the effectiveness of the investigated encoders:
The performance of a featureless learner (**FL** condition) was estimated as a conservative baseline for each dataset.
In regression problems, FL predicts the mean of the target variable in the training set for each observation in the test set.
In classification problems, the most frequent class of the target within the training set is predicted.
For each dataset, we also consider a random forest model without encoding (**none** condition), to compare the use of encoding methods in a random forest with a natural categorical splitting approach.
We use the random forest in ranger [@wright_2017] which provides efficient categorical feature support by ordering levels once before starting the tree growing algorithm [@wright_2019].
In the **remove** high cardinality features control condition, we remove features with a high number of levels above some threshold and use one-hot encoding (without collapsing rare levels) for the remaining features.
This condition reflects on whether including high cardinality features does indeed improve predictive performance.
Otherwise, the best encoding might just provide the least impairment compared to not including any high cardinality features.
We include an overview of available implementations in widely used machine learning frameworks for R and python in the Supplementary Material.

@@

Table \ref{tab:datasets} summarizes the datasets used in our benchmark study.
We specifically investigate datasets that contain categorical variables with a large number of levels, including many well-known datasets
from previous studies [@cerda_2018; @prokhorenkova_2018; @coors_2018; @kuhn_2019].
All datasets can be downloaded from the `OpenML` platform [@vanschoren_2013] based on the name or the displayed OmlId.
The datasets include `r nrow(descr_dat[Classes == 0])` regression, `r nrow(descr_dat[Classes == 2])` binary classification, and `r nrow(descr_dat[Classes > 2])` multiclass classification problems (between `r descr_dat[Classes > 2][, min(Classes)]` and `r descr_dat[Classes > 2][, max(Classes)]` classes).
To assess the imbalance in categorical variables we computed the normalized entropy for each categorical variable. The maximum normalized entropy is $1$, which corresponds to a uniform distribution, while a lower number indicates a larger imbalance.
Sample sizes range between `r descr_dat[, min(Obs)]` and `r descr_dat[, max(Obs)]` observations.
The total number of features ranges between `r descr_dat[, min(NumFeats + BinFeats + CatFeats)]` and `r descr_dat[, max(NumFeats + BinFeats + CatFeats)]`.
Datasets contain between `r descr_dat[, min(sapply(strsplit(HighCardLevels, " "), length))]` and `r descr_dat[, max(sapply(strsplit(HighCardLevels, " "), length))]` categorical features with more than $10$ levels, with the maximum number of levels for one feature in a dataset ranging between `r descr_dat[, min(MaxCatLvls)]` and `r descr_dat[, max(MaxCatLvls)]`.
Missing values are present in about half of the datasets.

@@

<!--
A relevant question we aim to answer is what the minimum number of levels is, for which more involved encoding methods should be used. WE CANNOT REALLY ANSWER THIS WITH ONLY 3 DIFFERENT HCT VALUES...
-->
It is widely assumed that more involved encoding methods are only advantageous for variables with a high number of levels, while simple indicator encoding is more appropriate for a small number of levels.
To make our benchmark results more realistic, a high cardinality threshold (HCT) parameter with values of $10$, $25$, and $125$ was introduced which determined varying configurations for the different encoders.
For indicator encoding, the $HCT - 1$ most frequent levels are encoded together with a single collapsed category for the remaining levels.
For integer, frequency, hash, leaf, impact, and glmm encoders, only features with more than HCT levels in the training set were encoded with the respective strategy, while the remaining categorical features were one-hot encoded.
HCT is used as the hash size in hash encoding.
In the remove control condition, features with more than HCT levels in the training set are removed from the feature set.
Based on the number of categorical features and levels per feature, some HCT settings were removed from the benchmark for some combinations of dataset $\times$ encoder.
This made sure that encoders always affect at least one feature and that the remove condition always removes at least one feature.
If several HCT settings of an encoder would lead to identical encoding strategies for all features of a dataset, we only kept the condition with the smallest HCT value.

>>>>

abstract: |
  Because most machine learning (ML) algorithms are designed for numerical inputs, efficiently encoding categorical variables is a crucial aspect during data analysis. An often encountered problem are high cardinality features, i.e. unordered categorical predictor variables with a high number of levels. We study techniques that yield numeric representations of categorical variables which can then be used in subsequent ML applications. We focus on the impact of those techniques on a subsequent algorithm's predictive performance, and -- if possible -- derive best practices on when to use which technique. We conducted a large-scale benchmark experiment, where we compared different encoding strategies together with five ML algorithms (lasso, random forest, gradient boosting, k-nearest neighbors, support vector machine) using datasets from regression, binary- and multiclass- classification settings. Throughout our study, regularized versions of target encoding (i.e. using target predictions based on the feature levels in the training set as a new numerical feature) consistently provided the best results. Traditional encodings that make unreasonable assumptions to map levels to integers (e.g. integer encoding) or to reduce the number of levels (possibly based on target information, e.g. leaf encoding) before creating binary indicator variables (one-hot or dummy encoding) were not as effective.
keywords:
- supervised machine learning
- benchmark
- high-cardinality categorical features
- target encoding
- dummy encoding
- generalized linear mixed models

@@

*Target-based* methods try to incorporate information about the target values associated with a given level.
Early strategies aimed to reduce the number of levels by methods like hierarchical clustering or decision trees based on statistics of the target variable, although this has been rarely described in the scientific literature [for a brief mention, see @micci_barreca_2001].
The basic idea of more advanced methods called target, impact, mean, or likelihood encoding is to use the training set to make a simple prediction of the target for each level of the categorical feature, and to use the prediction as the numerical feature value $\hat{x}_l$ for the respective level.
An early formal description of this strategy is @micci_barreca_2001.
In simple target encoding for regression problems, the mean target value in the training set from all observations with a certain feature level is used as the numeric value to encode that level for all observations: $\hat{x}_l = \frac{\sum_{i:x^{train}_i = l}y^{train}_i}{N_l}$.
Simple target encoding often does not perform well with rare levels, where it tends to overfit to the training data and fails to generalize well for new observations.
In the extreme case of a categorical feature with unique values (e.g. some hashed ID variable) studied in @prokhorenkova_2018, the mean target for each level of this feature in simple target encoding is similar to the true target value of a single observation.
Based on the encoded feature, all observations can be predicted perfectly in the training set, even if the original variable did not contain any useful information.
ML models would place a high priority on such an encoded feature during training but would perform badly on test data.
To avoid this, practitioners often use regularized target encoding with a smoothing parameter that shrinks those effects towards the global mean [@micci_barreca_2001].
An alternative strategy is to combine target encoding with cross-validation (CV) techniques [@prokhorenkova_2018].

@@

Several small-scale studies have been previously conducted.
Those works did not yield conclusive results due to narrower scopes or not considering high cardinality variables.
One benchmark ($6$ datasets) on encoding high cardinality features (cardinality between $103$ and $9095$) in combination with gradient boosting has been published on the Kaggle forums [@prokopev_2018].
Different versions of target encoding are compared with indicator, integer, and frequency encoding.
They recommend combining the smoothed version of target encoding with 4- or 5-fold CV, never using simple target encoding and using indicator encoding only for small datasets.
Interestingly, frequency encoding did perform well in many cases.
@coors_2018 performed a benchmark ($12$ datasets) on encoding high cardinality features (maximum number of levels per dataset between 10 and 25847) when developing the automatic gradient boosting (autoxgboost) library [@thomas_2018]. They compared different variants of target encoding with integer and indicator encoding. In their benchmarks, target encoding only improved over target-agnostic methods on $2$ datasets, while it led to worse results on 4 datasets. For a smaller number of levels, indicator and integer encoding yielded similar results.
As those studies only consider gradient boosting and a limited amount of datasets, it is unclear whether results generalize to other datasets and ML algorithms.
Several other publications studied encoding text data based on similarity [@cerda_2018; @cerda_2020] and employed indicator or target encoding as baselines.
Another line of work studies variable encodings employed within specific ML models.
@wright_2019 study treatments for categorical variables in random forests together with dummy and integer encoding (18 datasets, cardinality between $3$ and $38$), concluding that indicator and integer encoding perform subpar in comparison to methods that re-order levels according to the target variable.
@prokhorenkova_2018 compared their new CatBoost variant of target encoding to smoothed target encoding without CV, hold-out, and leave-one-out CV on $8$ datasets.
While the CatBoost method performed best, hold-out came second and target encoding without CV performed worst.
An overview of encoding techniques tailored towards neural networks (e.g. the widely adopted *entity embeddings* by @guo_2016) is provided in @hancock.
They present a survey of indicator- and embedding-based methods but no benchmark study.
In contrast, our work is the first to focus on high cardinality variables and techniques that are agnostic to the subsequent ML method.
We study this problem on a larger variety of datasets and settings.
\newline

The main goal of our study is to assess the impact different categorical encoding techniques have on subsequent models' predictive performance.
As the optimal encoding might differ depending on the ML algorithm, we consider various state-of-the-art algorithms, *regularized linear models* (LASSO), *random forests* (RF), *gradient tree boosting* (GB), *k-nearest neighbors* (KNN), and *support vector machines* (SVM).
To find default settings for high cardinality features, we analyze a variety of datasets with different characteristics including regression, binary classification, and multiclass classification problems.
Because our methods vary in runtime and complexity, we are interested whether more complex methods are to be preferred, or if simpler approaches suffice.
We also study the relationship between a feature's cardinality and the choice of encoding technique by varying the minimum number of levels above which features are transformed.

@@

We survey a broad set of categorical encoding techniques and conduct a comprehensive benchmark study with a focus on high cardinality features.
We study $7$ different encoding techniques in conjunction with $5$ commonly used ML algorithms across $24$ datasets, both from a classification and a regression regime. 
We give a detailed description of our study design to highlight important considerations for studying high cardinality features.
Our results provide an overview of the performance of various approaches heavily used in the literature.
After a discussion of results concerning predictive performance, we provide further analyses seeking to inform practitioners which methods to apply.
This includes an important discussion of runtimes.

@@

Pseudocode is presented in the main text (for non-standard encodings), or the Supplementary Material (for standard methods).
An important detail is how encoding techniques treat new levels during the prediction phase.

@@

The simplest strategy for categorical features is integer encoding (also called ordinal encoding).
We randomly map the observed levels from the training set to the integers $1$ to $L$.
Although new levels could be mapped to $L+1$ or $0$, model predictions would be arbitrary as the integer order does not carry information.
Thus, we encode new levels as missing values and use mode imputation to obtain the integer which matches the most frequent level in the training set.
Integer encoding should only be an acceptable strategy for tree-based models, which can separate all original levels with repeated splits.

@@

Frequency encoding maps each level to its observed frequency in the training set ($\hat{x}_l = N_l$).
This assumes a functional relationship between the frequency of a level and the target.
It implicitly reduces the number of levels, and the subsequent model can best differentiate between levels with dissimilar frequencies.
This approach is heavily used in natural language processing to encode token or n-gram counts.
We encode new levels with a frequency of $1$.

@@

We use indicator encoding as an umbrella term for two common strategies to encode categorical features with a small to moderate number of levels: one-hot and dummy encoding.
One-hot encoding transforms the original feature into $L$ binary indicator columns, each representing one original level.
An observation is coded with $1$ for the indicator column representing its level ($x^{train}_i = l$) and $0$ for all other indicators.
Dummy encoding results in only $L-1$ indicator columns.
A reference feature level is chosen that is encoded with $0$ in all indicator columns.
For one-hot encoding, the zero vector can be used to encode new levels which were not observed during training.
For dummy encoding, it is not useful to collapse new levels to the often arbitrary reference category; in our case the first level in alphabetical order.
We replace new levels in the prediction phase with the most frequent level in the training set.
In datasets with a high total number of feature levels, constructing all indicator variables can be detrimental to the predictive performance or at least increase the computational load.
As indicator encoding is practically infeasible for high cardinality variables (a feature with $10^5$ levels would add $10^5$ new columns), we limit the number of indicator coded features by collapsing rare levels beyond a varying threshold to a single *other* category before encoding.

@@

Hash Encoding can be used to compute indicator variables based on a hash function [@weinberger_2009].
The basic idea is to transform each feature level $l$ into an integer $hash(l) \in \mathbb{N}$, based on its label.
This integer is then transformed into an indicator representation, with $1$ in indicator column number $(hash(l) \mod hash.size) + 1$ and $0$ in all remaining columns [@kuhn_2019].
Some levels are hashed to the same indicator representation.
The smaller the $hash.size$, the higher the number of collapsed levels.
The number of indicators is often effectively lower than $hash.size$, as some indicators can be constant in the training set (we remove those columns for both training and prediction).
Although we could jointly hash multiple features, we hash each feature separately to improve comparability with the other encoders.

@@

Leaf encoding fits a decision tree on the training set to predict the target based on the categorical feature.
Each level is encoded by the number of the terminal node, in which an observation with the respective level ends up.
In that way, leaf encoding combines feature levels with similar target values.
We use the rpart package in R [@therneau_2018] which grows CARTs with categorical feature support that can be pruned based on internal performance estimates from 10-fold CV.
Thus, our leaf encoder automatically uses an "optimal" number of new levels.
To speed up the computation for multiclass classification, our implementation uses the ordering approach presented in @wright_2019.
New levels are encoded with the arbitrary number of the terminal node with most observations during training.
Encoded values are treated as a new categorical feature and encoded by one-hot encoding.
Our leaf encoder can be thought of as a simplification of the approach suggested by @grabczewski_2003.

@@

An early formal description of a so-called impact, target or James-Stein encoder was provided by @micci_barreca_2001.
The basic idea is to encode each feature level with the conditional target mean (regression) or the conditional relative frequency of one or more target classes (classification).
The impact encoder for classification uses a logit link and transforms the original feature into $C$ numeric features, each representing one target class.
A smoothing parameter $\epsilon$ is introduced to avoid division by zero.
This parameter could be used to further regularize towards the unconditional mean.
We always choose a small $\epsilon = 0.0001$ as we want to compare simple target encoding with the regularized encoder introduced next.
Weights of evidence encoding from the credit scoring classification literature [@hand_1997] is almost identical to impact encoding, but without regularization or centering.

@@

Smoothed target encoding [@micci_barreca_2001] can be interpreted as a simple (generalized) linear mixed model (glmm) in which the target is predicted by a random intercept for each feature level in addition to a fixed global intercept.
This connection is described in @kuhn_2019.
To achieve regularized impact encoding, we implemented glmm encoders for regression, binary, and multiclass classification which are described in Algorithm \ref{alg:glmm-regr} (regression) and in the Supplementary Material (classification).
The encoded value for each level is based on the spherical conditional mode estimates.
In regression, the conditional modes are similar to the mean target value for each level, weighted by the relative observed frequency of that level in the training set [@gelman_2006].
For technical details, see @bates_2018.
Additionally, the estimate of the fixed intercept can be used during the prediction phase to encode new feature levels that were not observed in the training set.
In multiclass classification, we fit $C$ one vs. rest glmms resulting in one encoded feature per class.
An important advantage of using a glmm over impact encoding with a smoothing parameter is that a reasonable amount of regularization is determined automatically and tuning the complete ML pipeline is not necessary.

A further strategy to avoid overfitting in target encoding is to combine it with CV to train the encoder on independent observations without limiting the data to train the ML model.
We provide an implementation which combines target encoding based on glmms with CV.
During the training phase, we partition the data using CV into $n.folds$ and fit a glmm on each resulting training set.
For each observation, there is exactly one glmm that did not use that observation for model fitting and can be safely used for encoding.
Note that the $n.folds$ CV models (for $n.folds$ > 1) are only used during the training phase.
In the prediction phase, feature values are always encoded by a single glmm fitted to the complete training set.
We study this method in three different settings: without CV (noCV), with $5-$ (5CV) and with $10-$ fold CV (10CV).
In our study, we use the lmer (regression) and glmer (classification) functions from the lme4 package in R [@bates_2015] as an efficient way to fit glmms.
\begin{algorithm}
  \caption{GLMM Encoding Regression}
  \label{alg:glmm-regr}
  \begin{algorithmic}
    \State \underline{Training:} require $n.folds \in \mathbb{N}$
     \State fit simple random intercept model: $y^{train}_i = \beta_{0l} + \epsilon_{i} = \gamma_{oo} + u_l + \epsilon_{i}$ on $\mathcal{D}^{train}$
     \State with $u_l \overset{iid}{\sim} N(0,\tau^2)$, $\epsilon_i \overset{iid}{\sim} N(0,\sigma^2)$ and $x^{train}_i = l$, $l \in \mathcal{L}^{train}$
    \If{$n.folds = 1$}
      \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}
        \State $\hat{x}^{train}_i = \hat{\beta}_{0l}^{\mathcal{D}^{train}}$ with $x^{train}_i = l$
      \EndFor
    \Else\ {use $n.folds$ cross-validation scheme to make training sets $\mathcal{D}^{train}_1, \dots, \mathcal{D}^{train}_{n.folds}$}
      \State and fit simple random intercept model on each $\mathcal{D}^{train}_{m}$
      \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}
        \State $\hat{x}^{train}_i = \hat{\beta}_{0l}^{\mathcal{D}^{train}}$ with $x^{train}_i = l$ based on the model m
        \State with $(x^{train}_i, y^{train}_i) \notin \mathcal{D}^{train}_m$
      \EndFor
    \EndIf
    \State \underline{Prediction:}
    \For{$x^{new}$}
      \If{$x^{new} \in \mathcal{L}^{train}$}
        \State $\hat{x}^{new} = \hat{\beta}_{0l}^{\mathcal{D}^{train}}$ with $x^{new} = l$ based on full model fitted on $\mathcal{D}^{train}$
      \Else\ {$\hat{x}^{new} = \hat{\gamma}_{00}$ based on full model fitted on $\mathcal{D}^{train}$}
      \EndIf
    \EndFor
  \end{algorithmic}
\end{algorithm}

@@

We include three control conditions to better understand the effectiveness of the investigated encoders:
The performance of a featureless learner (**FL** condition) was estimated as a conservative baseline for each dataset.
In regression problems, FL predicts the mean of the target variable in the training set for each observation in the test set.
In classification problems, the most frequent class of the target within the training set is predicted.
For each dataset, we also consider a RF without encoding (**none** condition), to compare the use of encoding methods with a natural categorical splitting approach.
The ranger [@wright_2017] implementation provides efficient categorical feature support by ordering levels once before starting the tree growing algorithm [@wright_2019].
In the **remove** high cardinality features control condition, we omit features with a high number of levels above some threshold and use one-hot encoding (without collapsing rare levels) for the remaining features.
This condition reflects on whether including high cardinality features does indeed improve predictive performance.
Otherwise, the best encoding might just provide the least impairment compared to not including any high cardinality features.
We include an overview of available implementations in widely used ML frameworks for R and python in the Supplementary Material.

@@

Table \ref{tab:datasets} summarizes our benchmark datasets.
We specifically investigate datasets that contain categorical variables with a large number of levels, including many well-known datasets
from previous studies [@cerda_2018; @prokhorenkova_2018; @coors_2018; @kuhn_2019].
All datasets can be downloaded from the `OpenML` platform [@vanschoren_2013] based on the name or the displayed OmlId.
The datasets include `r nrow(descr_dat[Classes == 0])` regression, `r nrow(descr_dat[Classes == 2])` binary classification, and `r nrow(descr_dat[Classes > 2])` multiclass classification problems (between `r descr_dat[Classes > 2][, min(Classes)]` and `r descr_dat[Classes > 2][, max(Classes)]` classes).
To assess the imbalance in categorical variables we computed the normalized entropy for each categorical variable.
The maximum normalized entropy is $1$, which corresponds to a uniform distribution, while a lower number indicates a larger imbalance.
Sample sizes range between `r descr_dat[, min(Obs)]` and `r descr_dat[, max(Obs)]` observations.
The total number of features ranges between `r descr_dat[, min(NumFeats + BinFeats + CatFeats)]` and `r descr_dat[, max(NumFeats + BinFeats + CatFeats)]`.
Datasets contain between `r descr_dat[, min(sapply(strsplit(HighCardLevels, " "), length))]` and `r descr_dat[, max(sapply(strsplit(HighCardLevels, " "), length))]` categorical features with more than $10$ levels, with the maximum number of levels for one feature in a dataset ranging between `r descr_dat[, min(MaxCatLvls)]` and `r descr_dat[, max(MaxCatLvls)]`.
Missing values are present in about half of the datasets.

@@

<!--
A relevant question we aim to answer is what the minimum number of levels is, for which more involved encoding methods should be used. WE CANNOT REALLY ANSWER THIS WITH ONLY 3 DIFFERENT HCT VALUES...
-->
It is often assumed that advanced encoding methods are only advantageous for variables with a high number of levels, while simple indicator encoding is more appropriate for a small number of levels.
To make our benchmark results more realistic, a high cardinality threshold (HCT) parameter with values of $10$, $25$, and $125$ was introduced which determined varying configurations for the different encoders.
For indicator encoding, the $HCT - 1$ most frequent levels are encoded together with a single collapsed category for the remaining levels.
For integer, frequency, hash, leaf, impact, and glmm encoders, only features with more than HCT levels in the training set were encoded with the respective strategy, while the remaining categorical features were one-hot encoded.
HCT is used as the hash size in hash encoding.
In the remove control condition, features with more than HCT levels in the training set are removed from the feature set.\footnote{Based on the number of categorical features and levels per feature, some HCT settings were removed from the benchmark for some combinations of dataset $\times$ encoder.
This made sure that encoders always affect at least one feature and that the remove condition always removes at least one feature.
If several HCT settings of an encoder would lead to identical encoding strategies for all features of a dataset, we only kept the condition with the smallest HCT value.}
